We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-21 paper/paper.tex
edit

The spatial distribution of satellite galaxies around our Milky Way
(MW) and the M31 galaxy is becoming a stringent test for structure
formation theories in a explicit cosmological context. 
This started with the suggestion of a Magellanic Plane, a flattened
structure of satellite galaxies around the MW, first suggested by
\cite{1976RGOB..182..241K} and \cite{1976MNRAS.174..695L}.

\cite{2005A&A...431..517K} then quantified that the
highly planar distribution of the 11 classical MW satellites has less
than a $0.5\%$ chance  to happen by chance if the
parent distribution is spherically symmetric, interpreting this as a
challenge to the Lambda Cold Dark Matter (LCDM) paradigm.
The same year it was recognized, using numerical simulations, that the
satellites should not be compared with an spherically
symmetric distribution given that Dark Matter halos in LCDM
are expected to be triaxial
\citep{2005MNRAS.363..146L, 2005ApJ...629..219Z}.
Nevertheless, estimates of the chances to find simulated satellites as
planar as in the MW were inconclusive. 
\cite{2005MNRAS.363..146L} reported that
satellite planes should be expected in every DM while
\cite{2005ApJ...629..219Z} gave an estimate of a $2\%$ probability to
have an arrangement as extreme as in observations.

Later, \cite{2007MNRAS.374.1125M} used simulations to confirm the low
chances ($<0.5\%$) found in \cite{2005A&A...431..517K}.
This result was challenged by \cite{2009MNRAS.399..550L} who continue
to report high chances to find a planar configuration;
two more recent numerical experiment with high resolution simulations
(\cite{2013MNRAS.429..725S} and \cite{2016MNRAS.457.1931S}) report
contradicting results (low and high chances to find the observational
result, respectively) but precise figures are not quoted in either of
these three reports. \citep{2015ApJ...815...19P,2016MNRAS.457.1931S}.
Meanwhile, \cite{2013MNRAS.429.1502W} and \cite{2014ApJ...789L..24P} reported
on new tests using high resolution simulations and cosmological
volumes arriving at a chance of $6\%$ and $0.77\%$, respectively, to
have a satellite distribution as triaxial as the observations.

In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All published studies agree on the point that the global satellite distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.    
Table \ref{table:M31} and Table \ref{table:MW} summarize some more
details on the results we have just mentioned.
Those tables also include the results from this paper using the
methodolgy we describe in the upcoming sections.

@@

Trying to reconcile and understand the seemly conflicting or inconclusive
results on the MW has its origin on the frequentist fashion generally used to
compute probabilities.
Usually, the process starts by building a high level parent sample in the
simulation and then counting how many elements has a subset
meeting some criteria. 
The first inconvenient with this approach is that the probability
estimate is made against whatever turns out to be typical in each
simulation. 
Comparing results across simulations would require first
characterizing the simulations at the high level parent samples.
The second inconvenient is that it is hard to find systems that fully resemble
the LG, this means that the high level parent sample has a small size
making it hard to find elements that fit some criteria.

In this paper we surmount the first difficulty by setting as direct 
point of reference an spherical satellite distribution and not the
simulations themselves.
The spherical distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We then characterize each satellite distribution in terms of the
scalars describing its deviation from the spherical distribution.
We handle the second difficulty by building an explicit analytic
probability distribution of the deviations from asphericity; this
solves the problem of having a common reference point to compare
simulations. 
We then use different simulations to estimate the parameters in this
distribution to later use it as a parent sample to generate any
desired number of samples that are by construction statistically
compatible with simulations, overcoming the problem of having a small
number of systems in the parent simulations.

>>>>

The spatial distribution of satellite galaxies around our Milky Way
(MW) and the M31 galaxy is becoming a stringent test for structure
formation theories in a explicit cosmological context.

This started with the suggestion of a Magellanic Plane, a flattened
structure of satellite galaxies around the MW, first suggested by
\cite{1976RGOB..182..241K} and \cite{1976MNRAS.174..695L}.  
\cite{2005A&A...431..517K} then quantified that the
highly planar distribution of the 11 classical MW satellites has less
than a $0.5\%$ chance  to happen by chance if the
parent distribution is spherically symmetric, interpreting this as a
challenge to the Lambda Cold Dark Matter (LCDM) paradigm.
The same year it was recognized, using numerical simulations, that
this comparison was unfair given given that Dark Matter halos in LCDM
are expected to be triaxial and not spherical
Nevertheless, some estimates of the chances to find simulated satellites as
planar as in the MW 
were as low as $2\%$ \citep{2005ApJ...629..219Z} while others expected
planar satellite configurations in every DM halo
\citep{2005MNRAS.363..146L}.

Later, \cite{2007MNRAS.374.1125M} used simulations to confirm the low
chances ($<0.5\%$) found in \cite{2005A&A...431..517K}.
This result was challenged by \cite{2009MNRAS.399..550L} who continue
to report high chances to find a planar configuration;
two more recent numerical experiment with high resolution simulations
(\cite{2013MNRAS.429..725S} and \cite{2016MNRAS.457.1931S}) report
contradicting results (low and high chances to find the observational
result, respectively) but precise figures are not quoted in either of
these three reports. 
Meanwhile, \cite{2013MNRAS.429.1502W} and \cite{2014ApJ...789L..24P} reported
on new tests using high resolution simulations and cosmological
volumes arriving at a chance of $6\%$ and $0.77\%$, respectively, to
have a satellite distribution as triaxial as the observations.

In the case of M31, observational studies have found a planar satellite
distribution in a special subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All published studies agree on the point that the spatial distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.    
Table \ref{table:M31} and Table \ref{table:MW} summarize some more
details on the results we have just mentioned \footnote{
Those tables also include the results from this paper using the
methodolgy we describe in the upcoming sections.}.

@@

Trying to reconcile and understand the seemly conflicting or inconclusive
results on the MW has its origin on the frequentist fashion generally used to
compute probabilities.
Usually, the process starts by building a high level parent sample in the
simulation and then counting how many elements has a subset
meeting some criteria. 
The first inconvenient with this approach is that the probability
estimate is made against whatever turns out to be typical in each
simulation. 
Comparing results across simulations would require first
characterizing all the simulations at the high level parent samples,
something that is difficult to do in practice.
The second inconvenient is that it is hard to find systems that fully resemble
the LG, this means that the high level parent sample has a small size
making it hard to find elements that fit some criteria.

In this paper we control the first effect by setting as direct 
point of reference an spherical satellite distribution and not the
simulations themselves.
The spherical satellite distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We characterize each satellite distribution in terms of the
scalars describing its deviation from the spherical distribution.
We then build an explicit analytic
probability distribution of the deviations from asphericity; this
solves the problem of having a common reference point to compare
simulations. 
We then use different simulations to estimate the parameters in this
distribution to later use it as a parent sample to generate any
desired number of samples that are by construction statistically
compatible with simulations, overcoming the problem of having a small
number of systems in the parent simulations.
